Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2025-10-08 Manuscript/endo_herbarium_GCB.tex
commiting accepted version of manuscipt, but still need to organize figure files
organized figure captions and updated data statement

@@

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = \linewidth]{../Plots/collections_map.png}
	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple). \firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:map}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{../Plots/year_plot.png}
	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species. (B)\firstrevise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects. Colored points are binned means of the observed endophyte presence/absence data (black dashes). Colors represent each host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple) and point size represents the number of specimens.}
	\label{fig:temporal}
\end{figure}

@@

\subsection*{How spatially variable are temporal trends in endophyte prevalence?}
While there was an overall increase in endophyte prevalence, our model revealed hotspots and coldspots of change across the host species' ranges, which are mapped in Fig. \ref{fig:svc_time_map} across geographic ranges predicted by MaxEnt species distribution models.
In some regions, posterior mean estimates of spatially varying temporal trends indicate that \emph{A. hyemalis} and \emph{A. perennans} experienced increases in prevalence by as much as $2$\% per year over the study period. \secondrevise{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for the \emph{Agrostis} species. (Fig. \ref{fig:svc_time_map_CI})}
\secondrevise{In contrast}, \emph{E. virginicus} experienced increases up to around $1$\% per year, \secondrevise{with uncertainty ranging between $3.5$\% increases and $2.5$\% decreases (Fig. \ref{fig:svc_time_map_CI})}
\secondrevise{Taken together, both} \emph{Agrostis} species show areas of both strong increasing and declining prevalence, while \emph{E. virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
Notably, endophytes are predicted to have increased most strongly towards the western range edge of \emph{A. hyemalis} (Fig. \ref{fig:svc_time_map}A) and across the northeastern US for \emph{A. perennans} (Fig. \ref{fig:svc_time_map}B). \linelabel{R1C7-begin}
\firstrevise{Broad increases in prevalence on average, along with increases towards range edges that had low historic prevalence result in range expansions of the symbiosis for both \emph{Agrostis} species (Fig. \ref{fig:svc_space_year_map}). 
Increases in prevalence were strongest in regions with low historic prevalence for the \emph{Agrostis} species (Fig. \ref{fig:initialprev_trend_plot} A-B), but for \emph{E. virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}
\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species ((A)\emph{A. hyemalis}; (B)\emph{A. perennans}; (C)\emph{E. virginicus}). \firstrevise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence. \firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:svc_time_map}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020. \firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for (A, D) \emph{A. hyemalis}, (B, E) \emph{A. perennans}, and (C, F) \emph{E. virginicus}. Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:svc_space_year_map}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \firstrevise{estimated from the \emph{post hoc} climate regression analysis}. Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple). Slopes with greater than 95\% \firstrevise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed. Points are the \firstrevise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \firstrevise{for the \emph{post hoc} climate regression analysis}.}
	\label{fig:climate_correlation_plot}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
		 The \firstrevise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys. The model captured regional trends across (A) longitude and (B) latitude. Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines: \emph{A. hyemalis}: green, orange, \emph{E. virginicus}: purple) from the herbarium model. Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval). (C) Comparison of \firstrevise{contemporary} observed \firstrevise{population prevalence} vs. predicted endophyte prevalence shows that contemporary test data had more variance between populations than \firstrevise{in model} predictions.}
	\label{fig:contemptestplot}
\end{figure}

@@

	\section*{Data and Code Availability}
	\sloppy{
	Data from this publication can be found through a publicly available repository (https://doi.org/10.5061/dryad.rn8pk0pn0).
	Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that will be permanently archived upon publication.
}
\newpage{}
	\bibliographystyle{plainnat}
	\bibliography{endo_herbarium}

@@

	\section*{Supporting Methods}
	\subsection*{ODMAP Protocol}\label{sec:sdm}
{\color{blue}{Overview}}\\
\textbf {Model  purpose}: Mapping current distribution of \emph{Epichloë} host species. \\
\textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
\textbf {Study area}: Eastern North America \\
%Scale of analysis
\textbf {Spatial extent}: -125.0208, -66.47917, 24.0625, 49.9375 (xmin, xmax, ymin, ymax).\\
\textbf {Spatial resolution}: 0.04166667, 0.04166667 (x, y).\\
\textbf {Temporal extent}: 1990 to 2020.\\
\textbf {Boundary}: Natural.\\
{\color{blue}{Data}}\\
\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America. We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
\textbf{Response data type}: occurrence record, presence-only.\\
\textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
\textbf{Climatic data}:  raster data extracted from PRISM;  30-year normal mean and standard deviation of temperature and of precipitation for three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December).  \\
{\color{blue}{Model }}\\
\textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
\textbf{Algorithms}: Maximum entropy (maxent)\\
\textbf{Workflow}: We  described the workflow in the methods section of the manuscript. \\
\textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium \\
\textbf{Data availability}: Data was accessed through open-source R packages \emph{rgbif}. \emph{A. hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, \emph{A. perennans} \cite{https://doi.org/10.15468/dl.wtne24}, \emph{E.  virginicus} \cite{https://doi.org/10.15468/dl.yt76uh} \\
{\color{blue}{Assessment}}\\
We used AUC to test model performance.\\
{\color{blue}{Prediction }}\\
We predicted the probability of presence of the host species as a binary maps (presence or absence)

>>>>

	\noindent{} $\ast$ Corresponding author; e-mail:jcf221@miami.edu

@@

	\textit{Running Title:} Increasing prevalence of \emph{Epichloë} endophytes

@@

@@

@@

\subsection*{How spatially variable are temporal trends in endophyte prevalence?}
While there was an overall increase in endophyte prevalence, our model revealed hotspots and coldspots of change across the host species' ranges, which are mapped in Fig. \ref{fig:svc_time_map} across geographic ranges predicted by MaxEnt species distribution models.
In some regions, posterior mean estimates of spatially varying temporal trends indicate that \emph{A. hyemalis} and \emph{A. perennans} experienced increases in prevalence by as much as $2$\% per year over the study period. \secondrevise{Posterior estimates of uncertainty in spatially varying slopes indicate that these hotspots of change may have experienced increases of up to $5$\% per year while declines in prevalence may be as great as $-4$\% per year for the \emph{Agrostis} species. (Fig. \ref{fig:svc_time_map_CI})}
\secondrevise{In contrast}, \emph{E. virginicus} experienced increases up to around $1$\% per year, \secondrevise{with uncertainty ranging between $3.5$\% increases and $2.5$\% decreases (Fig. \ref{fig:svc_time_map_CI})}
\secondrevise{Taken together, both} \emph{Agrostis} species show areas of both strong increasing and declining prevalence, while \emph{E. virginicus} had an overall weaker and geographically more homogeneous increase in endophyte prevalence.
Notably, endophytes are predicted to have increased most strongly towards the western range edge of \emph{A. hyemalis} (Fig. \ref{fig:svc_time_map}A) and across the northeastern US for \emph{A. perennans} (Fig. \ref{fig:svc_time_map}B). \linelabel{R1C7-begin}
\firstrevise{Broad increases in prevalence on average, along with increases towards range edges that had low historic prevalence result in range expansions of the symbiosis for both \emph{Agrostis} species (Fig. \ref{fig:svc_space_year_map}). 
Increases in prevalence were strongest in regions with low historic prevalence for the \emph{Agrostis} species (Fig. \ref{fig:initialprev_trend_plot} A-B), but for \emph{E. virginicus} trends did not differ according to historic prevalence (\ref{fig:initialprev_trend_plot} C).} \linelabel{R1C7-end}

@@

@@

@@

	%\section*{Data and Code Availability}
	%\sloppy{
	%Data from this publication can be found through a publicly available repository (https://doi.org/10.5061/dryad.rn8pk0pn0).
	%Code for analyses can be found through a publicly available repository (https://github.com/joshuacfowler/EndoHerbarium) that is %permanently archived on Zenodo: DOI: 10.5281/zenodo.17298101. Code includes methods for downloading publically available %datasets, including temperature and precipitation data from the PRISM climate group (https://prism.oregonstate.edu) and GBIF %occurrence records for \emph{Epichloë} host species (\emph{A. hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, \emph{A. %perennans} \cite{https://doi.org/10.15468/dl.wtne24}, \emph{E.  virginicus} \cite{https://doi.org/10.15468/dl.yt76uh}).
%}
\newpage{}
	\bibliographystyle{plainnat}
	\bibliography{endo_herbarium}

\newpage{}
\section*{Figure Captions}

\begin{figure}[h]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/collections_map.png}
	\caption[Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes]{\textbf{Collection locations of herbarium specimens sampled for \emph{Epichloë} endophytes.} Specimens span eastern North America from nine herbaria, and are colored by host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple). \firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:map}
\end{figure}

\begin{figure}[H]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/year_plot.png}
	\caption[Temporal trends in endophyte prevalence.]{\textbf{Temporal trends in endophyte prevalence.} (A) Histograms show the frequency of scored specimens through time for each host species. (B)\firstrevise{ Lines show mean endophyte prevalence predicted by the endophyte prevalence model} over the study period along with the 50\%  and 95\% CI bands incorporating parameter uncertainty and variation associated with collector and scorer random effects. Colored points are binned means of the observed endophyte presence/absence data (black dashes). Colors represent each host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple) and point size represents the number of specimens.}
	\label{fig:temporal}
\end{figure}

\begin{figure}[H]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/svc_time_map.png}
	\caption[Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species.]{Predicted posterior mean of spatially-varying slopes representing change in endophyte prevalence for each host species ((A)\emph{A. hyemalis}; (B)\emph{A. perennans}; (C)\emph{E. virginicus}). \firstrevise{Spatially-varying trends are estimated from the endophyte prevalence model.} Color indicates the relative change in predicted endophyte prevalence. \firstrevise{Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:svc_time_map}
\end{figure}

\begin{figure}[H]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/svc_space_map_year.png}
	\caption[Predicted endophyte prevalence for each host species in 1895 and 2020.]{\firstrevise{Predicted endophyte prevalence for each host species in 1895 and 2020. \firstrevise{Predictions of prevalence come from the endophyte prevalence model.} Color indicates the posterior mean endophyte prevalence for (A, D) \emph{A. hyemalis}, (B, E) \emph{A. perennans}, and (C, F) \emph{E. virginicus}. Map lines delineate study areas and do not necessarily depict accepted national boundaries.}}
	\label{fig:svc_space_year_map}
\end{figure}

\begin{figure}[H]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/climate_trends_plot_intercept.png}
	\caption[Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.]{\textbf{Relationships between predicted trends in endophyte prevalence and changes in seasonal climate drivers.} Lines show marginal predicted relationship between spatially-varying trends in endophyte prevalence and changes in mean and variability of climate ((A): mean temperature, (B): cumulative precipitation, (C): standard deviation in temperature, (D): standard deviation in precipitation) \firstrevise{estimated from the \emph{post hoc} climate regression analysis}. Confidence bands represent the 50 and 95\% CI,  colored by host species (\emph{A. hyemalis}: green, \emph{A. perennans}: orange, \emph{E. virginicus}: purple). Slopes with greater than 95\% \firstrevise{posterior} probability of being either positive or negative are represented as solid lines while those that have less than 95\% probability are dashed. Points are the \firstrevise{values of pre-computed SVC trends  and climate drivers at} 250 randomly sampled pixels across each host's distribution used in model fitting \firstrevise{for the \emph{post hoc} climate regression analysis}.}
	\label{fig:climate_correlation_plot}
\end{figure}

\begin{figure}[H]
	\centering
	%\includegraphics[width = \linewidth]{../Plots/contemp_test_plot.png}
	\caption[Predictive performance for contemporary test data.]{\textbf{Predictive performance for contemporary test data.} (A)
		The \firstrevise{endophyte prevalence model}, trained on historic herbarium collection data, performed modestly at predicting prevalence in contemporary population surveys. The model captured regional trends across (A) longitude and (B) latitude. Crosses indicate predicted mean prevalence along with the 95\% CI (colored lines: \emph{A. hyemalis}: green, orange, \emph{E. virginicus}: purple) from the herbarium model. Contemporary prevalence is represented by grey points (point size reflects sample size) along with trend lines from generalized linear models (black line and shaded 95\% confidence interval). (C) Comparison of \firstrevise{contemporary} observed \firstrevise{population prevalence} vs. predicted endophyte prevalence shows that contemporary test data had more variance between populations than \firstrevise{in model} predictions.}
	\label{fig:contemptestplot}
\end{figure}

@@

	\section*{Supporting Methods}
	\subsection*{ODMAP Protocol}\label{sec:sdm}
{\color{blue}{Overview}}\\
\textbf {Model  purpose}: Mapping current distribution of \emph{Epichloë} host species. \\
\textbf {Target species}: \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. \\
\textbf {Study area}: Eastern North America \\
%Scale of analysis
\textbf {Spatial extent}: -125.0208, -66.47917, 24.0625, 49.9375 (xmin, xmax, ymin, ymax).\\
\textbf {Spatial resolution}: 0.04166667, 0.04166667 (x, y).\\
\textbf {Temporal extent}: 1990 to 2020.\\
\textbf {Boundary}: Natural.\\
{\color{blue}{Data}}\\
\textbf {Observation type}: Occurrence records from  Global Biodiversity Information Facility and herbarium collection across eastern North America. We used 713 occurrences records for \emph{Agrostis hyemalis}, 656 occurrence records for \emph{Agrostis perennans} and 2338 for \emph{Elymus virginicus}.\\
\textbf{Response data type}: occurrence record, presence-only.\\
\textbf{Coordinate reference system}: WGS84 coordinate reference system (EPSG:4326 code)\\
\textbf{Climatic data}:  raster data extracted from PRISM;  30-year normal mean and standard deviation of temperature and of precipitation for three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December).  \\
{\color{blue}{Model }}\\
\textbf{Model assumption}: We assumed that the target species are at equilibrium with their environment. \\
\textbf{Algorithms}: Maximum entropy (maxent)\\
\textbf{Workflow}: We  described the workflow in the methods section of the manuscript. \\
\textbf{Software}:  All statistics were performed using Maxent 3.3.4 and R4.3.1 with packages terra, usdm, spThin and dismo.\\
\textbf{Code availability}: Available through this link: https://github.com/joshuacfowler/EndoHerbarium and archived on Zenodo (DOI: 10.5281/zenodo.17298101)\\
\textbf{Data availability}: Data was accessed through open-source R packages \emph{rgbif}. \emph{A. hyemalis} \cite{https://doi.org/10.15468/dl.ryqqcv}, \emph{A. perennans} \cite{https://doi.org/10.15468/dl.wtne24}, \emph{E.  virginicus} \cite{https://doi.org/10.15468/dl.yt76uh} \\
{\color{blue}{Assessment}}\\
We used AUC to test model performance.\\
{\color{blue}{Prediction }}\\
We predicted the probability of presence of the host species as a binary maps (presence or absence)
